Lists
- ordered collection of objects
- one-dimensional, hetergeneous
- ex: list of vectors, list of matrices, list of characters, list of functions, etc.
- indexing starts at 1, not 0

creating a list (list of employees with specific details)
```{r}
empId = c(1, 2, 3, 4) 
empName = c("Debi", "Sandeep", "Subham", "Shiba")
numberOfEmp = 4
 
empList = list(empId, empName, numberOfEmp)
 
print(empList)
```

naming list components
```{r}
my_named_list <- list(name = "Sudheer", age = 25, city = "Delhi")

print(my_named_list)
```

accessing list components
```{r}
empId = c(1, 2, 3, 4)
empName = c("Debi", "Sandeep", "Subham", "Shiba")
numberOfEmp = 4

empList = list(
  "ID" = empId,
  "Names" = empName,
  "Total Staff" = numberOfEmp
  )
print(empList)

cat("Accessing name components using $ command\n")
print(empList$Names)
```

access components by indices

```{r}
empId = c(1, 2, 3, 4)
empName = c("Debi", "Sandeep", "Subham", "Shiba")
numberOfEmp = 4

empList = list(
  "ID" = empId,
  "Names" = empName,
  "Total Staff" = numberOfEmp
)
print(empList)

cat("Accessing name components using indices\n")
print(empList[[2]])

cat("Accessing Sandeep from name using indices\n")
print(empList[[2]][2])

cat("Accessing 4 from ID using indices\n")
print(empList[[1]][4])
```

modifying components of a list

```{r}
empId = c(1, 2, 3, 4)
empName = c("Debi", "Sandeep", "Subham", "Shiba")
numberOfEmp = 4

empList = list(
  "ID" = empId,
  "Names" = empName,
  "Total Staff" = numberOfEmp
)
cat("Before modifying the list\n")
print(empList)

empList$`Total Staff` = 5
empList[[1]][5] = 5
empList[[2]][5] = "Kamala"

cat("After modified the list\n")
print(empList)
```

concatenation of lists

```{r}
empId = c(1, 2, 3, 4)
empName = c("Debi", "Sandeep", "Subham", "Shiba")
numberOfEmp = 4

empList = list(
  "ID" = empId,
  "Names" = empName,
  "Total Staff" = numberOfEmp
)
cat("Before concatenation of the new list\n")
print(empList)

empAge = c(34, 23, 18, 45)
empList = c(empName, empAge)

cat("After concatenation of the new list\n")
print(empList)
```

adding item to a list (append function)

```{r}
my_numbers = c(1,5,6,3)

append(my_numbers, 45)

my_numbers
```

deleting components of a list

```{r}
empId = c(1, 2, 3, 4)
empName = c("Debi", "Sandeep", "Subham", "Shiba")
numberOfEmp = 4

empList = list(
  "ID" = empId,
  "Names" = empName,
  "Total Staff" = numberOfEmp
)
cat("Before deletion the list is\n")
print(empList)

cat("After Deleting Total staff components\n")
print(empList[-3])

cat("After Deleting sandeep from name\n")
print(empList[[2]][-2])
```

merging lists (placing all lists into a single list)

```{r}
lst1 <- list(1,2,3)
lst2 <- list("Sun","Mon","Tue")

new_list <- c(lst1, lst2)

print(new_list)
```
