Matrixes
- two-dimenstional arrangement of data in rows and columns
- homogeneous

syntax:
matrix(data, nrow, ncol, byrow, dimnames)
- data: values you want entered
- nrow: # of rows
- ncol: # of columns
- byrow: FALSE by default, if TRUE, values will be assigned by rows
- dimnames: names of rows and columns

example matrix
```{r}
A = matrix(
  c(1, 2, 3, 4, 5, 6, 7, 8, 9), 
  nrow = 3,   
  ncol = 3,         
  byrow = TRUE          
)

rownames(A) = c("a", "b", "c")
colnames(A) = c("c", "d", "e")

cat("The 3x3 matrix:\n")
print(A)
```

accessing rows
```{r}
A = matrix(
  c(1, 2, 3, 4, 5, 6, 7, 8, 9), 
  nrow = 3,             
  ncol = 3,             
  byrow = TRUE          
)
cat("The 3x3 matrix:\n")
print(A)

cat("Accessing first and second row\n")
print(A[1:2, ])
```

accessing columns
```{r}
A = matrix(
  c(1, 2, 3, 4, 5, 6, 7, 8, 9), 
  nrow = 3,             
  ncol = 3,             
  byrow = TRUE          
)
cat("The 3x3 matrix:\n")
print(A)

cat("Accessing first and second column\n")
print(A[, 1:2])
```

accessing submatrices
```{r}
A = matrix(
  c(1, 2, 3, 4, 5, 6, 7, 8, 9), 
  nrow = 3,             
  ncol = 3,             
  byrow = TRUE          
)
cat("The 3x3 matrix:\n")
print(A)

cat("Accessing the first three rows and the first two columns\n")
print(A[1:3, 1:2])
```

rbind()
```{r}
A = matrix(
  c(1, 2, 3, 4, 5, 6, 7, 8, 9), 
  nrow = 3,             
  ncol = 3,             
  byrow = TRUE          
)
cat("The 3x3 matrix:\n")
print(A)

B = matrix(
  c(10, 11, 12),
  nrow = 1,
  ncol = 3
)
cat("The 1x3 matrix:\n")
print(B)

C = rbind(A, B)

cat("After concatenation of a row:\n")
print(C)
```

cbind()
```{r}
A = matrix(
  c(1, 2, 3, 4, 5, 6, 7, 8, 9), 
  nrow = 3,             
  ncol = 3,             
  byrow = TRUE          
)
cat("The 3x3 matrix:\n")
print(A)

B = matrix(
  c(10, 11, 12),
  nrow = 3,
  ncol = 1,
  byrow = TRUE
)
cat("The 3x1 matrix:\n")
print(B)

C = cbind(A, B)

cat("After concatenation of a column:\n")
print(C)
```

adding rows
```{r}
number <- matrix(
  c(1, 2, 3, 4, 5, 6, 7, 8, 9), 
  nrow = 3,             
  ncol = 3,             
  byrow = TRUE          
)
cat("Before inserting a new row:\n")
print(number)

new_row <- c(10, 11, 12)

A <- rbind(number[1, ], new_row, number[-1, ])

cat("\nAfter inserting a new row:\n")
print(A)
```

adding column
```{r}
number <- matrix(
  c(1, 2, 3, 4, 5, 6, 7, 8, 9), 
  nrow = 3,             
  ncol = 3,             
  byrow = TRUE          
)
cat("Before adding a new column:\n")
print(number)

new_column <- c(10, 11, 12)

number <- cbind(number, new_column)

cat("\nAfter adding a new column:\n")
print(number)
```

row deletion
```{r}
A = matrix(
  c(1, 2, 3, 4, 5, 6, 7, 8, 9), 
  nrow = 3,             
  ncol = 3,             
  byrow = TRUE          
)
cat("Before deleting the 2nd row\n")
print(A)

A = A[-2, ]

cat("After deleted the 2nd row\n")
print(A)
```

column deletion
```{r}
A = matrix(
  c(1, 2, 3, 4, 5, 6, 7, 8, 9), 
  nrow = 3,             
  ncol = 3,             
  byrow = TRUE          
)
cat("Before deleting the 2nd column\n")
print(A)

A = A[, -2]

cat("After deleted the 2nd column\n")
print(A)
```