---
title: "STAT 3080 — Function Reference & Examples"
author: "Study Guide"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
```

---

# Part 1: `dplyr` — Row Verbs

## `filter()` — Keep rows matching a condition

```{r}
# Comma between conditions = AND (&)
flights <- nycflights13::flights

flights |>
  filter(month == 1, day == 1) |>
  slice_head(n = 3)

# OR condition
flights |>
  filter(month == 1 | month == 2) |>
  slice_head(n = 3)
```

## `between()` — Shorthand for range filtering

```{r}
# between(x, a, b) is equivalent to x >= a & x <= b
flights |>
  filter(between(dep_delay, -5, 5)) |>
  slice_head(n = 3)
```

## `arrange()` — Sort rows

```{r}
# Default: ascending. desc() for descending.
flights |>
  arrange(desc(dep_delay)) |>
  slice_head(n = 3)

# Multiple columns: sorts by first, breaks ties with second
flights |>
  arrange(month, desc(dep_delay)) |>
  slice_head(n = 3)
```

## `slice()` — Select rows by index

```{r}
# Pick specific row numbers (can repeat for sampling with replacement!)
flights |>
  slice(c(1, 3, 3, 100)) |>
  select(month, day, dep_delay)
```

## `slice_head()` / `slice_tail()` — First or last n rows

```{r}
flights |> slice_head(n = 3) |> select(year, month, day)
flights |> slice_tail(n = 3) |> select(year, month, day)
```

## `slice_sample()` — Random rows

```{r}
set.seed(42)

# Without replacement (simple random sample — default)
flights |>
  slice_sample(n = 5, replace = FALSE) |>
  select(month, day, carrier)

# With replacement (bootstrap resample — needed for bootstrap!)
flights |>
  slice_sample(n = 5, replace = TRUE) |>
  select(month, day, carrier)
```

## `distinct()` — Unique rows

```{r}
# All unique combinations of origin and dest
flights |>
  distinct(origin, dest) |>
  slice_head(n = 5)

# .keep_all = TRUE retains all other columns (first occurrence)
flights |>
  distinct(origin, dest, .keep_all = TRUE) |>
  slice_head(n = 3) |>
  select(origin, dest, carrier, month)
```

---

# Part 2: `dplyr` — Column Verbs

## `select()` — Keep or drop columns

```{r}
# By name
flights |> select(year, month, day) |> slice_head(n = 3)

# Range
flights |> select(year:day) |> slice_head(n = 3)

# Exclude
flights |> select(!year:day) |> slice_head(n = 3)

# By type
flights |> select(where(is.character)) |> slice_head(n = 3)

# Pattern matching
flights |> select(starts_with("dep")) |> slice_head(n = 3)
flights |> select(ends_with("delay")) |> slice_head(n = 3)
flights |> select(contains("arr")) |> slice_head(n = 3)
```

## `rename()` — Rename a column (keeps all others)

```{r}
# new_name = old_name
flights |>
  rename(tail_num = tailnum) |>
  select(tail_num, carrier) |>
  slice_head(n = 3)
```

## `mutate()` — Add or modify columns

```{r}
# Add new columns (appear on the right by default)
flights |>
  mutate(
    gain      = dep_delay - arr_delay,
    speed_mph = distance / air_time * 60
  ) |>
  select(dep_delay, arr_delay, gain, speed_mph) |>
  slice_head(n = 4)

# .before and .after to control position
flights |>
  mutate(gain = dep_delay - arr_delay, .before = 1) |>
  slice_head(n = 3) |>
  select(1:5)

# .keep = "used" — only show columns involved in the calculation
flights |>
  mutate(
    gain  = dep_delay - arr_delay,
    hours = air_time / 60,
    .keep = "used"
  ) |>
  slice_head(n = 4)

# Set column to NULL to delete it
flights |>
  mutate(year = NULL) |>
  slice_head(n = 2) |>
  select(1:5)
```

## `if_else()` — Vectorized conditional (handles NA safely)

```{r}
# if_else(condition, value_if_TRUE, value_if_FALSE)
flights |>
  mutate(
    delay_status = if_else(dep_delay > 0, "Delayed", "On time")
  ) |>
  select(dep_delay, delay_status) |>
  slice_head(n = 5)
```

## `relocate()` — Move columns

```{r}
# Default: brings to the front
flights |>
  relocate(time_hour, air_time) |>
  slice_head(n = 2) |>
  select(1:5)

# .before / .after for precise control
flights |>
  relocate(starts_with("arr"), .before = dep_time) |>
  slice_head(n = 2) |>
  select(1:6)
```

---

# Part 3: `dplyr` — Group Verbs

## `group_by()` + `summarize()` — Group-level summaries

```{r}
# group_by() adds invisible grouping metadata
# summarize() collapses each group to one row
flights |>
  group_by(month) |>
  summarize(
    avg_delay = mean(dep_delay, na.rm = TRUE),
    n_flights  = n()
  ) |>
  slice_head(n = 4)
```

## `n()` — Count rows in current group

```{r}
# Only works inside summarize(), mutate(), filter() after group_by()
flights |>
  group_by(carrier) |>
  summarize(total = n()) |>
  arrange(desc(total)) |>
  slice_head(n = 5)
```

## `count()` — Quick frequency table

```{r}
# Equivalent to group_by() |> summarize(n = n())
flights |>
  count(carrier, origin) |>
  slice_head(n = 5)
```

## Multiple groups + `.groups` argument

```{r}
# When grouped by >1 variable, summarize() drops the LAST group
# Use .groups = "drop_last" to suppress the warning
flights |>
  group_by(month, carrier) |>
  summarize(avg_delay = mean(dep_delay, na.rm = TRUE),
            .groups = "drop_last") |>
  slice_head(n = 4)
```

## `ungroup()` — Remove grouping

```{r}
# After ungroup(), summarize() gives a single grand total
flights |>
  group_by(month) |>
  ungroup() |>
  summarize(grand_avg = mean(dep_delay, na.rm = TRUE))
```

## `.by =` inside `summarize()` — Alternative to `group_by()`

```{r}
# Equivalent result, but output is NOT a grouped tibble
flights |>
  summarize(
    avg_delay = mean(dep_delay, na.rm = TRUE),
    n = n(),
    .by = c(origin, dest)
  ) |>
  slice_head(n = 4)
```

## `slice_max()` / `slice_min()` within groups

```{r}
# Most delayed flight per destination
flights |>
  group_by(dest) |>
  slice_max(arr_delay, n = 1, with_ties = FALSE) |>
  select(dest, arr_delay, carrier) |>
  slice_head(n = 5)
```

---

# Part 4: Tidy Data — `tidyr`

## `pivot_longer()` — Wide → Long

```{r}
# billboard has columns wk1, wk2, ..., wk76
# Each week column is actually a VALUE of the "week" variable
billboard |>
  pivot_longer(
    cols         = starts_with("wk"),  # which columns to collapse
    names_to     = "week",             # new column for old column names
    values_to    = "rank",             # new column for cell values
    values_drop_na = TRUE              # drop NA rows (songs that fell off)
  ) |>
  slice_head(n = 5)
```

## `parse_number()` — Extract number from a string

```{r}
# After pivoting, "week" is character "wk1", "wk2", etc.
# parse_number() strips non-numeric characters
billboard |>
  pivot_longer(starts_with("wk"), names_to = "week", values_to = "rank",
               values_drop_na = TRUE) |>
  mutate(week = parse_number(week)) |>
  slice_head(n = 4)
```

## `pivot_wider()` — Long → Wide

```{r}
# cms_patient_experience has 6 rows per organization (one per measure)
# We want one row per organization with one column per measure
cms_patient_experience |>
  pivot_wider(
    id_cols     = starts_with("org"),  # uniquely identifies each observation
    names_from  = measure_cd,          # values of this column become new col names
    values_from = prf_rate             # fill new columns with these values
  ) |>
  slice_head(n = 3)
```

---

# Part 5: `ggplot2` — Building Plots Layer by Layer

## The basic template

```{r, fig.height=3}
# Step 1: data only — empty canvas
ggplot(data = diamonds)

# Step 2: add aesthetics — axes appear
ggplot(data = diamonds, mapping = aes(x = carat, y = price))
```

## Adding geoms

```{r, fig.height=3}
# geom_point() — scatter plot
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 0.1)   # alpha handles overplotting
```

## Plot-level vs. geom-level aesthetics

```{r, fig.height=3}
# color at PLOT level → inherited by BOTH geoms → 3 separate lines
ggplot(diamonds, aes(x = carat, y = price, color = cut)) +
  geom_point(alpha = 0.05) +
  geom_smooth(method = "lm", se = FALSE)

# color at GEOM level → only affects geom_point → 1 overall line
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(color = cut), alpha = 0.05) +
  geom_smooth(method = "lm", se = FALSE)
```

## `labs()` — Labels

```{r, fig.height=3}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(color = cut), alpha = 0.1) +
  labs(
    title    = "Diamond price vs. carat weight",
    subtitle = "Colored by cut quality",
    x        = "Carat weight",
    y        = "Price (USD)"
  )
```

## EDA Plot Types by Variable Combination

```{r, fig.height=3}
# Numeric alone → histogram
ggplot(diamonds, aes(x = carat)) +
  geom_histogram(binwidth = 0.1)

# Categorical alone → bar chart
ggplot(diamonds, aes(x = cut)) +
  geom_bar()

# Numeric + Categorical → boxplot (use fct_reorder to sort by median)
ggplot(diamonds, aes(x = fct_reorder(cut, price, median), y = price)) +
  geom_boxplot()

# Numeric + Numeric → scatter (use alpha for overplotting)
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 1/20)

# Categorical + Categorical → geom_count
ggplot(diamonds, aes(x = cut, y = color)) +
  geom_count()
```

---

# Part 6: Sampling in R

## Deterministic sampling

```{r}
die <- tibble(face = 1:6)

# By index — note row 3 appears twice (sampling WITH replacement manually)
die |> slice(c(1, 3, 3, 6))
```

## `sample()` — Base R random sampling

```{r}
set.seed(42)

# Without replacement (default) — just a permutation
sample(1:6, size = 3, replace = FALSE)

# With replacement — essential for bootstrap!
sample(1:6, size = 6, replace = TRUE)
```

## `row_number()` — Add index column for systematic sampling

```{r}
die |>
  mutate(row_index = row_number()) |>
  relocate(row_index)
```

## `seq()` — For systematic sampling

```{r}
# Start randomly, then take every 2nd element
start <- sample(1:2, size = 1)
die |> slice(seq(start, n(), by = 2))
```

## `replicate()` — Repeat an experiment M times

```{r}
set.seed(42)
M <- 1000

# Each call draws a sample of size 10 from the die and computes the median
# simplify = FALSE → returns a list; bind_rows() converts to tibble
sampling_dist <- replicate(M,
  die |>
    slice_sample(n = 10, replace = TRUE) |>
    summarize(med = median(face)),
  simplify = FALSE
) |>
  bind_rows()

# dim() of a replicate() result when simplify = TRUE (default for vectors):
# rows = length of each result, cols = M
# Here we used simplify = FALSE so we get a list → bind_rows gives a tibble
dim(sampling_dist)  # M rows × 1 column (the "med" column)
```

## Understanding `dim()` with `replicate()`

```{r}
set.seed(42)

# When each replicate returns a VECTOR of length k,
# replicate(..., simplify = TRUE) gives a MATRIX: k rows × M cols
mat <- replicate(5, sample(1:6, size = 3, replace = TRUE))
dim(mat)  # 3 rows × 5 cols

# Key exam pattern: replicate(M, sample(data)) where data has n elements
# → matrix with n rows × M cols
grades_exam1 <- c(68, 69, 71, 67, 78)
boot_samps <- replicate(100, sample(grades_exam1))
dim(boot_samps)  # 5 rows × 100 cols
```

---

# Part 7: Bootstrap Confidence Intervals

## The full bootstrap CI workflow

```{r}
set.seed(3080)

# Original sample
x <- c(68, 69, 71, 67, 78)
n <- length(x)
R <- 10000   # number of bootstrap resamples

# --- Point estimate ---
theta_hat <- mean(x)    # replace mean() with any estimator (median, IQR, etc.)
cat("Point estimate:", theta_hat, "\n")

# --- Generate bootstrap resamples and compute statistic for each ---
boot_stats <- replicate(R,
  mean(sample(x, size = n, replace = TRUE))  # MUST use replace = TRUE
)

# --- Method 1: Basic (Bias-Corrected) Bootstrap CI ---
# Formula: (2*theta_hat - q_{1-a/2}, 2*theta_hat - q_{a/2})
boot_errors  <- boot_stats - theta_hat
sorted_errors <- sort(boot_errors)
alpha <- 0.05

lower_idx <- R * (1 - alpha/2)   # e.g., 9750
upper_idx <- R * (alpha/2)       # e.g., 250

basic_ci <- theta_hat - sorted_errors[c(lower_idx, upper_idx)]
cat("Basic bootstrap 95% CI:", basic_ci, "\n")

# --- Method 2: Percentile Bootstrap CI ---
# Just take the alpha/2 and 1-alpha/2 quantiles of the bootstrap distribution
pct_ci <- quantile(boot_stats, c(alpha/2, 1 - alpha/2))
cat("Percentile bootstrap 95% CI:", pct_ci, "\n")
```

## Bootstrap bias estimate

```{r}
# Bias estimate = mean of bootstrap stats - original estimate
bias_est <- mean(boot_stats) - theta_hat
cat("Bootstrap bias estimate:", bias_est, "\n")

# An unbiased-corrected estimator would be: theta_hat - bias_est
```

## Key conceptual checks

```{r}
# ✅ Correct bootstrap: replace = TRUE, size = n
mean(sample(x, size = 5, replace = TRUE))

# ❌ Wrong: replace = FALSE — always gives same values, no variability
# mean(sample(x, size = 5, replace = FALSE))

# ❌ Wrong: no size= argument — defaults to full permutation, no replacement
# mean(sample(x))
```

---

# Part 8: Randomization (Permutation) Tests

## Two-sample randomization test — step by step

```{r}
set.seed(3080)

# Hypothetical GPA data: did or did not submit SAT scores
group_A <- c(3.75, 3.80, 4.00, 3.33)   # submitted SAT (nA = 4)
group_B <- c(3.00, 2.67, 3.75)          # did not submit (nB = 3)

nA <- length(group_A)
nB <- length(group_B)
n  <- nA + nB

# Step 1: Observed test statistic
t0 <- abs(mean(group_A) - mean(group_B))
cat("Observed |diff in means|:", round(t0, 4), "\n")

# Step 2-3: Generate B permutation resamples
B <- 10000
all_gpas <- c(group_A, group_B)  # pool everyone together

perm_stats <- replicate(B, {
  shuffled <- sample(all_gpas)           # randomly reassign labels
  new_A    <- shuffled[1:nA]             # first nA go to "group A"
  new_B    <- shuffled[(nA + 1):n]       # rest go to "group B"
  abs(mean(new_A) - mean(new_B))         # compute test statistic
})

# Step 4: Permutation p-value (add 1 to numerator and denominator)
p_value <- (sum(perm_stats >= t0) + 1) / (B + 1)
cat("Permutation p-value:", round(p_value, 4), "\n")

# Step 5: Visualize the permutation distribution
tibble(T = perm_stats) |>
  ggplot(aes(x = T)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = t0, color = "red", linewidth = 1) +
  labs(title = "Permutation distribution of |diff in means|",
       subtitle = paste("Observed t0 =", round(t0, 3),
                        " | p-value =", round(p_value, 3)),
       x = "Test statistic", y = "Count")
```

---

# Part 9: Monte Carlo — Estimating Type I Error

## The core idea: each test outcome is a Bernoulli trial

```{r}
set.seed(3080)

# Simulate Type I error rate for a z-test when data are NOT normal
# True population: chi-squared with 2 df (mean = 2, sd = 2)
# H0: mu = 2 (true!), so any rejection IS a Type I error

n     <- 8       # small sample — CLT may not apply well
alpha <- 0.05
K     <- 10000   # Monte Carlo sample size
sigma <- 2       # known population SD

# Each replicate: draw sample, apply z-test, record reject/not reject
I <- replicate(K, {
  x    <- rchisq(n, df = 2)          # sample from true distribution
  z    <- (mean(x) - 2) / (sigma / sqrt(n))  # z statistic
  reject <- abs(z) > qnorm(1 - alpha/2)       # two-sided test
  as.integer(reject)
})

# Empirical Type I error rate (should be ≈ alpha if CLT applies)
I_bar <- mean(I)
cat("Empirical Type I error rate:", round(I_bar, 4), "\n")
cat("Nominal alpha:", alpha, "\n")

# 99% CI for the true Type I error rate p
alpha_star <- 0.01
margin <- qnorm(1 - alpha_star/2) * sqrt(I_bar * (1 - I_bar) / K)
cat("99% CI for p:", round(I_bar - margin, 4), "to", round(I_bar + margin, 4), "\n")
```

---

# Quick Reference: Key Distinctions

```{r, echo=FALSE}
tribble(
  ~Concept,                    ~Description,
  "n (experimental)",          "Size of original sample; determines shape of sampling distribution",
  "M or K (Monte Carlo)",      "# of simulation replicates; controls precision of the MC estimate",
  "R (bootstrap resamples)",   "# of bootstrap draws; controls precision of bootstrap CI",
  "replace = TRUE",            "REQUIRED for bootstrap; sampling WITH replacement",
  "replace = FALSE",           "For simple random samples; each element can only appear once",
  "Basic bootstrap CI",        "(2*theta_hat - q_high, 2*theta_hat - q_low) — bias-corrected",
  "Percentile bootstrap CI",   "(q_low, q_high) of the bootstrap distribution — simpler",
  "Bootstrap bias",            "mean(boot_stats) - theta_hat",
  "Permutation p-value",       "(# resamples with T >= t0 + 1) / (B + 1)",
  "LLN justifies...",          "Monte Carlo simulation AND bootstrap AND empirical distributions"
) |>
  knitr::kable(caption = "Key Concepts Cheat Sheet")
```
